---
title: "Association analysis with genomic phenotypes with FRET"
author: "Jean Morrison"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Association analysis with genomic phenotypes with FRET}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval=FALSE)
```

## Analysis pipeline (in progress)

1. Format your data. You will need:
  - Files containing the genomic phenotype (one per chromosome). These files should have one row per base-pair. The first column should give base-pair position and each subsequent column should give the phenotype value for one sample. Lines should look something like this:
  
  84419286 0 0 3.6 1.15 0.95 0 0 3.5 7.2 0 0.8 0 0 1.15 0 4.3 0 2.13 0 0 0 1.5 0
  
The phenotype files should have a header. The name of the first column should be 'pos' and the name of each subsequent column should be a sample id that is also used in the trait file. 

  - A trait file. This file should have a header. The first column should be the sample name (corresponding to the header of the phenotype file). Subsequent columns should contain the trait and any covariates you wish to adjust for.
  
1. Choose s0, z0, and zmin. There are (a few?) steps for choosing these values:

  1. Calculate test statistics for some large chunk of data. For example, lets say we wish to use the 50kb stretch between positions 84400000 84900000 on chromosome 1. We run
  
```{r}
stats <- maxes_table(dat.file, pheno.file, s0=0, n.perm=0, range=c(84400000, 84900000),
                        pheno.transformation=NULL, trait=c("x"), covariates=c(),
                        bandwidth=50, out.file=NULL, chrom="chr1",
                        stat.type="huber", smoother="ksmooth_0")

```

Note that we run with n.perm = 0 (no permutations yet!).
This creates an object that contains the names of the data files supplied, a data frame of test statistics and a data frame of smoothed test statistics (we don't really need these now but they can be nice to plot). The arguments pheno.transformation, triat, and covariates are used to specify the regression model. stat.type can be "huber" or "lm". 

  1. Feed the resulting stats into the choose_s0 function:

```{r}
choose_s0(beta=stats$stats$Beta, sd=stats$stats$SD)
```

  1. Once you've chosen s0, choose zmin:
 
```{r}
choose_zmin(beta=stats$stats$Beta, sd=stats$stats$SD, s0, pos=stats$stats$pos, bandwidth=50, smoother="ksmooth_0", zmin_quantile=0.9)
``` 
  We recomoend setting z0 to 0.3*zmin.
  
1. Calculate test statistics and permutation test statistics. This is a lot of computing and you may wish to break it into chunks. Run maxes_table for each chunk (specify the chunk with the range argument). You'll want to use and output file to save the data rather than return it.

```{r}
maxes_table(dat.file, pheno.file, s0, n.perm, zmin, z0=0.3*zmin, 
              range=c(84400000, 84900000), bandwidth=50, chrom="chr1",
              pheno.transformation=NULL, trait=c("x"), covariates=c(),
              out.file="chunk1_maxes_table.RData",
              stat.type="huber", smoother="ksmooth_0")

```
